---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# blockr.core

<!-- badges: start -->
[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![status](https://github.com/cynkra/blockr.core/actions/workflows/ci.yaml/badge.svg)](https://github.com/cynkra/blockr.core/actions/workflows/ci.yaml)
[![coverage](https://codecov.io/gh/cynkra/blockr.core/graph/badge.svg?token=VoOPRU65KA)](https://codecov.io/gh/cynkra/blockr.core)
<!-- badges: end -->

## Overview

blockr is a framework for data analysis, using a web-based point and click user interface.
Use blockr to create web apps that perform visual programming, leveraging the power of R.

## What is a block?

A block is a fundamental unit of a data analysis workflow.
A single block performs a single tasks, such as loading or filtering or visualisation.
Blocks can be joined together using a DAG to create powerful data workflows.

## What is blockr.core

`blockr.core` is the entry point to blockr, providing the core functionality needed to get started.
It can be run as a stand-alone app, but is designed with extensibility in mind.
A rich API is provided to connect your own blocks and override the default behaviours (e.g., you can add your own GUI layer or seralisation).

## Installation

You can install the development version of blockr.core from [GitHub](https://github.com/) with:

``` r
pak::pak("cynkra/blockr.core")
```

## Get started

Spin up an app with a single block:

```{r eval = FALSE}
library(blockr.core)
serve(new_dataset_block("iris"))
```

For a block that requires input, additionally pass static data:

```{r, eval = FALSE}
serve(
  new_merge_block(by = "name"),
  data = list(x = datasets::BOD, y = datasets::ChickWeight)
)
```

Use the board object to manage and spin up an app with multiple blocks (more on boards, links, and stacks, later):

```{r, eval = FALSE}
serve(
  new_board(
    blocks = c(
      a = new_dataset_block("BOD"),
      b = new_dataset_block("ChickWeight"),
      c = new_merge_block("Time")
    ),
    links = c(
      ac = new_link("a", "c", "x"),
      bc = new_link("b", "c", "y")
    ),
    stacks = list(ac = c("a", "c"))
  )
)
```

## Creating blocks

**Note:** If you are not already familiar with [Shiny modules](https://shiny.posit.co/r/articles/improve/modules/), it is recommended you do so before continuing with this section.

Blockr is built on top of Shiny. At it's heart, a block is just a specialised Shiny module that returns two additional values:

- **An expression** (called "expr"), a reactive expression defining a blocks computation. This allows R code to be exported and  data analyses to be recreated, outside of blockr and a reactive context.
- **A state object** (called "state"), a list of reactive values tracking user inputs. This allows blocks to separate user-defined state from inputs specified by blocks.

Blocks consists of three elements:

1. A **UI function** to define the user interface. 
2. A **Server function** to handle reactive logic that returns an expression and state object.
3. A **Constructor function** wraps the UI and server to intialize the block state.

Now, let's go through each element, building up a block template as we go.

### UI function

A UI function in blockr is the same as a UI function in a Shiny module.
This means that:

- The UI function signature is expected to contain a single `id` argument, which can be used with `shiny::NS()` to construct namespaced IDs.
- A call to appropriate shiny UI functions is expected that return `shiny.tag` or `shiny.tag.list` objects, typically via the use of `shiny::tagList()` in the UI to list separate UI elements.

Let's start to build our block template:

```{r, eval = FALSE}
ui <- function(id) {
  tagList(
    # Wrap widgets in `tagList()`
    textInput(
      NS(id, "my_input") # Use `shiny::NS()` to construct namespaces
    )
  )
}

```

### Server function

As inputs, a server function takes an id and any additional inputs from other blocks (e.g., data).

A server function shold return as ouput a `moduleServer()` call, defining:

- **expr**: A quoted reactive expression representing the block’s computation.
- **state**: A list of reactive values tracking user selections.

```{r eval = FALSE}
function(id, ...) {
  moduleServer(id, function(input, output, session) {
    # Reactive logic goes here

    # Return a list with "expr" and "state"
    list(
      expr = reactive(quote(identity(...))),
      state = list()
    )
  })
}
```


### Constructor function

The constructor function wraps the UI and server functions and initializes the block.

It should expose as arguments anything the user might set via the UI to control the block state.
In other words, any inputs in the UI function, should have a corresponding argument in the constructor function.
On the other hand, inputs from other blocks should not be exposed as arguments, as these are dynamically provided in the server function.

The return value should be a call to `new_block()` (or if applicable a call to the more specific *virtual* constructors `new_data_block()`, `new_transform_block()`, etc.).

We can finish out block template:

```{r eval = FALSE}
example_constructor_function(ui_state = character())
{
  ui <- function(id) {
    tagList(textInput(NS(id, "ui_state")))
  }

  server <- function() {
    # Reactive logic goes here

    # Return "expr" and "state"
    list(
      expr = reactive(quote(identity(...))),
      state = list(reactive(ui_state()))
    )
  }

  # Return call to `new_block()`
  new_block()
}

```

### Example

Putting this all together, we can create a ...

# OLD README ---

## How to create a block

A block constructor should expose as arguments anything that defines its *state*, i.e. things the user might set via UI which are not derived from input data. For example in a `shiny::selectInput()`, where the `choices` are data columns, this should not be a constructor argument (as this value will be computed from the data), whereas the `selected` value should be exposed.

The constructor return value should be a call to `new_block()` (or if applicable a call to the more specific *virtual* constructors `new_data_block()`, `new_transform_block()`, etc.). Arguments `server` and `ui` both expect closures with a specific structure.

### Block `server`

- A signature with as many arguments as data inputs is expected: zero for a data block, one for a transform block such as a select block, two for a join block and the special argument `...args` for variadic blocks, such as an `rbind` block (all, excluding the required first argument `id`).
- The first argument `id` should be passed to the `shiny::moduleServer()` call.
- As return value a call to `shiny::moduleServer()` is expected, containing a `module` function that in turn returns a list with entries `expr` and `state`.
- The block expression is a quoted reactive value that contains user-supplied components and is updated whenever user values change. Data names should match between expressio and the top-level function arguments.
- Block state is defined by the `state` entry which again is a list of reactives. The set of returned values should match (both in count and names) that of the constructor signature.

The `server` component of an identity transform block could be formed as

```{r eval = FALSE}
function(id, data) {
  moduleServer(
    id,
    function(input, output, session) {
      list(
        expr = reactive(quote(identity(data))),
        state = list()
      )
    }
  )
}
```

Note that `data` is the same name in the expression and the top-level function signature.

### Block `ui`

- The function signature is expected to contain a single `id` argument, which can be used with `shiny::NS()` to construct namespaced IDs.
- A call to appropriate shiny UI functions is expected that return `shiny.tag` or `shiny.tag.list` objects.
- The initial evaluation of the `ui` function will be performed in the context of the constructor scope (i.e. if a value of name `xyz` is bound the the constructor scope or a parent thereof, this value will be passed as `xyz` argument).

The `ui` component of an identity transform block is trivial:

```{r eval = FALSE}
function(id) {
  tagList()
}
```

Putting this together, a dataset block could be constructed as

```{r}
new_dataset_block <- function(
  dataset = character(),
  package = "datasets",
  ...
) {
  envir <- as.environment(paste0("package:", package))

  choices <- ls(envir = envir)
  choices <- choices[
    vapply(mget(choices, envir = envir), is.data.frame, logical(1L))
  ]

  new_data_block(
    function(id) {
      moduleServer(
        id,
        function(input, output, session) {
          dat <- reactiveVal(dataset)

          observeEvent(input$dataset, dat(input$dataset))

          list(
            expr = reactive(
              bquote(
                `::`(.(pkg), .(dat)),
                list(pkg = package, dat = dat())
              )
            ),
            state = list(
              dataset = reactive(input$dataset),
              package = package
            )
          )
        }
      )
    },
    function(id) {
      selectInput(
        inputId = NS(id, "dataset"),
        label = "Dataset",
        choices = choices,
        selected = dataset
      )
    },
    class = "dataset_block",
    ...
  )
}
```

Note that both `server` and `ui` are closures and therfore may refer to names bound in the constructor scope (e.g. `dataset` and `choices`). The `state` entry returned by the server module contains the static value `package` and this is needed for ser/deser (alongside the dynamic value `dataset`).

The `expr` return value is here constructed via a call to `base::bquote()`. Many other options exsist, such as using tools offered by the rlang package or by passing a string to `base::parse()`, e.g.

```{r eval = FALSE}
list(
  expr = reactive(parse(text = paste0(package, "::", dataset))[[1]]),
  ...
)
```

The block can then be manipulated via UI as

```{r, eval = FALSE}
serve(new_dataset_block("iris"))
```

Another example is a `utils::head()` block, such as the one offered as `new_head_block()`:

```{r}
new_head_block <- function(n = 6L, ...) {
  new_transform_block(
    function(id, data) {
      moduleServer(
        id,
        function(input, output, session) {
          n_rows <- reactiveVal(n)

          observeEvent(input$n, n_rows(input$n))

          observeEvent(
            nrow(data()),
            updateNumericInput(
              inputId = "n",
              value = n_rows(),
              min = 1L,
              max = nrow(data())
            )
          )

          list(
            expr = reactive(
              bquote(utils::head(data, n = .(n)), list(n = n_rows()))
            ),
            state = list(
              n = n_rows
            )
          )
        }
      )
    },
    function(id) {
      tagList(
        numericInput(
          inputId = NS(id, "n"),
          label = "Number of rows",
          value = n,
          min = 1L
        )
      )
    },
    dat_val = function(data) {
      stopifnot(is.data.frame(data) || is.matrix(data))
    },
    class = "head_block",
    ...
  )
}
```

Here, the state return value is a list of length 1, containing the current value for the only constructor argument. The current value for `n` is represented by a `shiny::reactiveVal()`, `n_rows()`, which is initialized with the values `n` in the constructor scope and updated on every change to `input$n`. Furthermore the `max` value for the `shiny::numericInput()` field is updated on every change to the number of data rows.

A final `block` constructor argument of note might be `dat_val`, which is an optional function that can be passed which signals to the framework, if data passed to a block, can actually be processed by the block.

An app containing such a head block can be spun up as

```{r, eval = FALSE}
serve(new_head_block(n = 10L), list(data = mtcars))
```

For an example with multiple data inputs, refer to examples such as `new_merge_block()`. Such a binary block (with arguments `x` and `y`) can be explored in a standalone app (with nonsensical inputs) as

```{r, eval = FALSE}
serve(
  new_merge_block(by = "Time"),
  data = list(x = datasets::BOD, y = datasets::ChickWeight)
)
```

The `data` argument to `serve()` expects a list with names components that match the server function signature, i.e. `data` for `new_head_block()` and `x`, `y` for `new_merge_block()`. Such names can be chosen freely by the block implementer (with the exception of integer-valued names, such as `` `1` ``, `` `2` ``, etc. which are reserved as positional arguments in `...args`).
