---
title: "2. Create a block"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Create a block}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(blockr.core)
```

## Creating blocks

**Note:** If you are not already familiar with [Shiny modules](https://shiny.posit.co/r/articles/improve/modules/), it is recommended you familiarise yourself with them before continuing with this section.

`blockr.core` is built on top of Shiny. At it's heart, a __block__ is nothing more than a specialised Shiny __module__ that returns two additional values:

- **An expression** (called "expr"), a reactive expression defining a blocks computation. This allows R code to be exported and  data analyses to be recreated, outside of blockr and a reactive context.
- **A state object** (called "state"), a list of reactive values tracking user inputs. This allows blocks to separate user-defined state from inputs specified by blocks.

```{mermaid}
flowchart LR
  subgraph block[Block]
    subgraph ctor[constructor]
      block_ui[UI]
      subgraph block_server[server]
        blk_state[State]
        blk_expr[Expression]
      end
    end
  end
```

While blocks *return* the "expr" and "state" values, blocks *consist* of three elements:

1. A **UI function** to define the user interface. 
2. A **Server function** to handle reactive logic that returns an expression and state object.
3. A **Constructor function** wraps the UI and server to intialize the block state.

Now, let's go through each element, building up a block template as we go.

### UI function

A UI function in blockr is the same as a UI function in a Shiny module.
This means that:

- The UI function signature is expected to contain a single `id` argument, which can be used with `shiny::NS()` to construct namespaced IDs.
- A call to appropriate shiny UI functions is expected that return `shiny.tag` or `shiny.tag.list` objects, typically via the use of `shiny::tagList()` in the UI to list separate UI elements.

Let's start to build our block template:

```{r, eval = FALSE}
ui <- function(id) {
  tagList(
    # Wrap widgets in `tagList()`
    textInput(
      NS(id, "my_input") # Use `shiny::NS()` to construct namespaces
    )
  )
}
```

### Server function

As inputs, a server function takes an id and any additional inputs from other blocks (e.g., data). For example, for the blocks shipped with blockr.core, this would result in zero inputs for a __data__ block, one for a __transform__ block such as a select block, two for a __join__ block and the special argument `...args` for __variadic__ blocks, such as an `rbind` block.

```{mermaid}
flowchart TB
  data_blk[data block 1]
  data_blk_2[data block 2]
  data_blk_3[data block 3]
  data_blk_4[data block 4]
  select_blk[select block]
  join_blk[join block]
  rbind_blk[rbind block]
  data_blk --> |data| select_blk
  data_blk_2 -->|data1| join_blk
  data_blk_3 --> |data2| join_blk
  
  data_blk --> |1| rbind_blk
  select_blk --> |2| rbind_blk
  data_blk_4 --> |3| rbind_blk
  join_blk --> |4| rbind_blk
```

A server function should return as ouput a `moduleServer()` call, defining:

- **expr**: A quoted reactive expression representing the block’s computation. Data names should match between expression and the top-level server function arguments.
- **state**: A list of reactive values tracking user selections. The set of returned values from the __state__ should match (both in count and names) that of the constructor signature.

```{r eval = FALSE}
server <- function(id, data) {
  moduleServer(id, function(input, output, session) {
    # Reactive logic goes here

    # Return a list with "expr" and "state"
    list(
      expr = reactive(quote(identity(data))),
      state = list(
        input_one = reactive(input_one()),
        input_two = reactive(input_two())
      )
    )
  })
}
```

Many options exist to build expressions, such as using tools offered by the `rlang` package or by passing a string to `base::parse()`.
As a block developer you are free to choose the tools that your prefer, as long as the evaluated expression returns non-reactive R code which can be executed outside of blockr.

The state is needed for __serialisation__ and __deserialisation__ to respectively save and restore the state of a blockr application.

### Constructor function

The __constructor__ function wraps the UI and server functions and initializes the block. Note that both `server` and `ui` are [closures](https://adv-r.hadley.nz/environments.html#function-environments) and therefore may refer to names bound in the constructor scope (e.g. `dataset` and `choices`).

It should expose as arguments anything the user might set via the UI to control the block __state__.
In other words, any inputs in the UI function, should have a corresponding argument in the constructor function.
On the other hand, inputs from other blocks should not be exposed as arguments, as these are dynamically provided in the server function.

The return value should be a call to `new_block()` (or if applicable a call to the more specific *virtual* constructors `new_data_block()`, `new_transform_block()`, etc.).

We can finish our block template:

```{r eval = FALSE}
example_constructor <- function(ui_state = character(), ...) {
  ui <- function(id) { #nolint
    tagList(textInput(NS(id, "ui_state")))
  }

  server <- function(id, data) {
    moduleServer(id, function(input, output, session) {
      # Reactive logic goes here

      # Return a list with "expr" and "state"
      list(
        expr = reactive(quote(identity(data))),
        state = list(
          # name must match what is defined in the constructor signature
          ui_state = <STATE_VALUE>
        )
      )
    })
  }

  # Return call to `new_block()`
  new_block(
    server = server,
    ui = ui,
    class = "my_block",
    ...
  )
}
```

The `...` is forwarded to `new_block` or any `new_*_block`.

### Example

Putting this all together, we can create a `utils::head()` block, such as the one offered as `new_head_block()`:

```{r}
new_head_block <- function(n = 6L, ...) {
  new_transform_block(
    function(id, data) {
      moduleServer(
        id,
        function(input, output, session) {
          n_rows <- reactiveVal(n)

          observeEvent(input$n, n_rows(input$n))

          observeEvent(
            nrow(data()),
            updateNumericInput(
              inputId = "n",
              value = n_rows(),
              min = 1L,
              max = nrow(data())
            )
          )

          list(
            expr = reactive(
              bquote(utils::head(data, n = .(n)), list(n = n_rows()))
            ),
            state = list(
              n = n_rows
            )
          )
        }
      )
    },
    function(id) {
      tagList(
        numericInput(
          inputId = NS(id, "n"),
          label = "Number of rows",
          value = n,
          min = 1L
        )
      )
    },
    dat_val = function(data) {
      stopifnot(is.data.frame(data) || is.matrix(data))
    },
    class = "head_block",
    ...
  )
}
```

Here, the state return value is a list of length 1, containing the current value for the only constructor argument. The current value for `n` is represented by a `shiny::reactiveVal()`, `n_rows()`, which is initialized with the values `n` in the constructor scope and updated on every change to `input$n`. Furthermore the `max` value for the `shiny::numericInput()` field is updated on every change to the number of data rows.

A final `block` constructor argument of note might be `dat_val`, which is an optional function that can be passed which signals to the framework, if data passed to a block, can actually be processed by the block.

An app containing such a head block can be spun up as

```{r, eval = FALSE}
serve(new_head_block(n = 10L), list(data = mtcars))
```

For an example with multiple data inputs, refer to examples such as `new_merge_block()`. Such a binary block (with arguments `x` and `y`) can be explored in a standalone app (with nonsensical inputs) as

```{r, eval = FALSE}
serve(
  new_merge_block(by = "Time"),
  data = list(x = datasets::BOD, y = datasets::ChickWeight)
)
```

The `data` argument to `serve()` expects a list with names components that match the server function signature, i.e. `data` for `new_head_block()` and `x`, `y` for `new_merge_block()`. Such names can be chosen freely by the block implementer (with the exception of integer-valued names, such as `` `1` ``, `` `2` ``, etc. which are reserved as positional arguments in `...args`).
